---
title: "Posterior Probability Calculation using LDA"
author:
  - name: David N Brown, Bill H Diplas
    affiliation: Memorial Sloan Kettering Cancer Center
output:
  html_document:
    theme: cerulean
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

This notebook fits a per-gene linear discriminant analysis (LDA) model on the
HPV-16 ctDNA training set and applies it to all HPV-16 samples to compute
posterior probabilities. The results are written to `../res/Posterior_Probability_ALL.txt`
and a QC summary is rendered below.

```{r setup, eval = TRUE, echo = FALSE, results = 'hide', include = FALSE}
suppressPackageStartupMessages({
  library("readr")
  library("dplyr")
  library("magrittr")
  library("reshape2")
  library("MASS")
  library("knitr")
})

# --- Data paths ---
url_manifest                 <- "../data/manifest.txt"
url_preanalytical_conditions <- "../data/pre-analytical_conditions.txt"
url_mrd_summary              <- "../data/TM-2021-MSK-002_MRD_Results_Summary_20220825.txt"
url_mutation_summary         <- "../data/mutation_summary.maf"
url_hpv_type                 <- "../data/TM-2021-MSK-002_HPV_Typing_WES_WGS.txt"
url_no_node_dissection       <- "../data/patients_0_nodal_dissection.txt"
url_bed_file                 <- "../data/hpv_31_33_35H_16_18.bed"
url_insert_metrics_by_gene   <- "../data/insert_metrics_by_gene.txt"

url_output <- "../res/Posterior_Probability_ALL.txt"

target_contigs <- c(
  "HPV-16" = "NC001526.4",
  "HPV-18" = "NC001357.1",
  "HPV-31" = "J04353.1",
  "HPV-33" = "M12732.1",
  "HPV-35" = "X74477.1"
)

contig_to_hpv <- c(
  "J04353.1"   = "HPV-31",
  "M12732.1"   = "HPV-33",
  "NC001357.1" = "HPV-18",
  "NC001526.4" = "HPV-16",
  "X74477.1"   = "HPV-35"
)

target_names <- c(
  "E1_1:1950", "E2_1891:2989", "E5_2985:3237",
  "E6_7124:7601", "E7_7603:7900", "L1_4774:6292"
)
```

```{r load_shared_data, eval = TRUE, include = FALSE}

manifest_raw <- readr::read_tsv(
  url_manifest, col_names = TRUE, col_types = cols(.default = col_character())
) %>%
  readr::type_convert() %>%
  dplyr::filter(!is.na(bam_file_name_hpv)) %>%
  dplyr::mutate(sample_uuid = paste0(sample_id_mskcc, "-", sample_id_invitae))

preanalytical_conditions <- readr::read_tsv(
  url_preanalytical_conditions, col_names = TRUE, col_types = cols(.default = col_character())
) %>%
  readr::type_convert()

mrd_smry <- readr::read_tsv(
  url_mrd_summary, col_names = TRUE, col_types = cols(.default = col_character())
) %>%
  readr::type_convert() %>%
  dplyr::mutate(
    sample_uuid = gsub("EP-D1-D1", "", Sample_ID_Archer, fixed = TRUE),
    sample_name = gsub("EP-D1-D1", "", Sample_ID_Archer, fixed = TRUE)
  )

hpv_smry <- readr::read_tsv(
  url_hpv_type, col_names = TRUE, col_types = cols(.default = col_character())
) %>%
  readr::type_convert() %>%
  dplyr::mutate(
    patient_id_mskcc = gsub("-T", "", patient_name, fixed = TRUE),
    patient_id_mskcc = gsub("155salv", "155", patient_id_mskcc, fixed = TRUE),
    hpv_type_wes     = gsub("HPV", "HPV-", hpv_type_wes, fixed = TRUE),
    hpv_type_wgs     = gsub("HPV", "HPV-", hpv_type_wgs, fixed = TRUE),
    hpv_type_wes_wgs = gsub("HPV", "HPV-", hpv_type_wes_wgs, fixed = TRUE),
    hpv_type_wes_wgs = dplyr::coalesce(hpv_type_wes_wgs, "Unknown")
  )

nodal_dissection_smry <- readr::read_tsv(
  url_no_node_dissection, col_names = TRUE, col_types = cols(.default = col_character())
) %>%
  readr::type_convert() %>%
  dplyr::rename(patient_id_mskcc = sample)

manifest <- manifest_raw %>%
  dplyr::left_join(preanalytical_conditions, by = "sample_id_mskcc") %>%
  dplyr::mutate(
    patient_id_mskcc = dplyr::case_when(
      is.na(patient_id_mskcc) & sample_id_mskcc == "21-144-03654" ~ "CTMS-164",
      TRUE ~ patient_id_mskcc
    ),
    sample_name = paste0(sample_id_mskcc, "-", sample_id_invitae)
  ) %>%
  dplyr::left_join(hpv_smry, by = "patient_id_mskcc")

mutation_smry <- readr::read_tsv(
  url_mutation_summary, col_names = TRUE, col_types = cols(.default = col_character())
) %>%
  readr::type_convert()

bed_file <- readr::read_tsv(
  url_bed_file, col_names = FALSE, col_types = cols(.default = col_character())
) %>%
  readr::type_convert() %>%
  dplyr::rename(chromosome = X1, start = X2, end = X3, gene_name = X4) %>%
  dplyr::select(-X5) %>%
  dplyr::filter(gene_name != "E1^E4")

insert_size_metrics <- readr::read_tsv(
  url_insert_metrics_by_gene, col_names = TRUE, col_types = cols(.default = col_character())
) %>%
  readr::type_convert() %>%
  dplyr::rename(sample_name = SAMPLE_NAME) %>%
  dplyr::filter(TARGET_REGION != "NC001526.4:3372-4794")

# --- Per-sample mutation AF summary ---
mut_af_smry <- mutation_smry %>%
  dplyr::filter(FILTER == "PASS") %>%
  dplyr::group_by(Tumor_Sample_Barcode) %>%
  dplyr::summarize(
    mean_af   = mean(t_maf),
    median_af = median(t_maf),
    max_af    = max(t_maf),
    .groups   = "drop"
  ) %>%
  dplyr::rename(sample_name = Tumor_Sample_Barcode)

# --- ctDNA-positive training set ---
smry_t_pos <- mut_af_smry %>%
  dplyr::filter(mean_af > 0.05) %>%
  dplyr::left_join(manifest, by = "sample_name") %>%
  dplyr::filter(hpv_type_wes_wgs %in% names(target_contigs)) %>%
  dplyr::left_join(
    dplyr::select(mrd_smry, sample_name, `MRD-Landmark_Result`),
    by = "sample_name"
  ) %>%
  dplyr::filter(`MRD-Landmark_Result` == "PRESENT") %>%
  dplyr::mutate(Is_ctDNA = "+ve")

# --- ctDNA-negative training set ---
neg_candidates <- mut_af_smry %>%
  dplyr::filter(mean_af < 0.0001, max_af < 0.001) %>%
  dplyr::left_join(manifest, by = "sample_name") %>%
  dplyr::left_join(nodal_dissection_smry, by = "patient_id_mskcc") %>%
  dplyr::filter(nd_event == 0, timepoint_days_since_start_of_RT > 730)

latest_per_patient <- neg_candidates %>%
  dplyr::group_by(patient_id_mskcc) %>%
  dplyr::summarize(
    timepoint_days_since_start_of_RT = max(timepoint_days_since_start_of_RT),
    .groups = "drop"
  )

smry_t_neg <- neg_candidates %>%
  dplyr::inner_join(latest_per_patient, by = c("patient_id_mskcc", "timepoint_days_since_start_of_RT")) %>%
  dplyr::left_join(
    dplyr::select(mrd_smry, sample_name, `MRD-Landmark_Result`),
    by = "sample_name"
  ) %>%
  dplyr::filter(`MRD-Landmark_Result` == "ABSENT") %>%
  dplyr::mutate(Is_ctDNA = "-ve")

shared_cols <- intersect(colnames(smry_t_pos), colnames(smry_t_neg))
smry_ft <- dplyr::bind_rows(
  dplyr::select(smry_t_pos, all_of(shared_cols)),
  dplyr::select(smry_t_neg, all_of(shared_cols))
) %>%
  dplyr::select(sample_name, Is_ctDNA)
```

```{r build_matrices, eval = TRUE, include = FALSE}

# --- Helper: build insert-size / read-count base from insert_size_metrics ---
build_smry_base <- function(filter_to_training = FALSE) {
  base <- insert_size_metrics %>%
    dplyr::left_join(mut_af_smry, by = "sample_name") %>%
    dplyr::left_join(manifest, by = "sample_name") %>%
    dplyr::mutate(hpv_type_wes_wgs = dplyr::coalesce(hpv_type_wes_wgs, "Unknown")) %>%
    dplyr::mutate(
      chromosome = vapply(TARGET_REGION, function(x) strsplit(x, ":", fixed = TRUE)[[1]][1], character(1)),
      start      = vapply(TARGET_REGION, function(x) strsplit(x, ":|-", perl = TRUE)[[1]][2], character(1)),
      end        = vapply(TARGET_REGION, function(x) strsplit(x, ":|-", perl = TRUE)[[1]][3], character(1))
    ) %>%
    readr::type_convert() %>%
    dplyr::left_join(bed_file, by = c("chromosome", "start", "end")) %>%
    dplyr::mutate(
      chromosome    = dplyr::recode(chromosome, !!!contig_to_hpv),
      mean_af       = dplyr::if_else(mean_af   == 0, 1e-6, mean_af),
      median_af     = dplyr::if_else(median_af == 0, 1e-6, median_af),
      max_af        = dplyr::if_else(max_af    == 0, 1e-6, max_af),
      start_end     = paste0(start, ":", end),
      aligned_reads = log10(READ_PAIRS + 1),
      insert_size   = log10(MEAN_INSERT_SIZE + 1)
    ) %>%
    dplyr::filter(chromosome == "HPV-16", hpv_type_wes_wgs == "HPV-16")

  if (filter_to_training) {
    base <- base %>%
      dplyr::left_join(dplyr::select(smry_ft, sample_name, Is_ctDNA), by = "sample_name") %>%
      dplyr::filter(!is.na(Is_ctDNA)) %>%
      dplyr::mutate(
        Is_ctDNA = dplyr::case_when(
          hpv_type_wes_wgs == "HPV-16" & Is_ctDNA == "+ve" ~ "+ve",
          TRUE ~ "-ve"
        ),
        Is_ctDNA = factor(Is_ctDNA, levels = c("+ve", "-ve"), ordered = TRUE)
      )
  }
  base
}

# --- Helper: pivot base into wide read-count + insert-size feature matrix ---
pivot_to_matrix <- function(base, has_label = FALSE) {
  id_vars <- if (has_label) c("sample_name", "Is_ctDNA") else "sample_name"

  reads_wide <- reshape2::dcast(
    base,
    formula   = as.formula(paste(paste(id_vars, collapse = " + "), "~ gene_name + start_end")),
    value.var = "aligned_reads",
    fill      = 0
  ) %>%
    dplyr::rename(
      `E1_1:1950_READ_COUNT`    = `E1_1:1950`,
      `E2_1891:2989_READ_COUNT` = `E2_1891:2989`,
      `E5_2985:3237_READ_COUNT` = `E5_2985:3237`,
      `E6_7124:7601_READ_COUNT` = `E6_7124:7601`,
      `E7_7603:7900_READ_COUNT` = `E7_7603:7900`,
      `L1_4774:6292_READ_COUNT` = `L1_4774:6292`
    )

  ins_wide <- reshape2::dcast(
    base,
    formula   = as.formula(paste(paste(id_vars, collapse = " + "), "~ gene_name + start_end")),
    value.var = "insert_size",
    fill      = 0
  ) %>%
    dplyr::rename(
      `E1_1:1950_INSERT_SIZE`    = `E1_1:1950`,
      `E2_1891:2989_INSERT_SIZE` = `E2_1891:2989`,
      `E5_2985:3237_INSERT_SIZE` = `E5_2985:3237`,
      `E6_7124:7601_INSERT_SIZE` = `E6_7124:7601`,
      `E7_7603:7900_INSERT_SIZE` = `E7_7603:7900`,
      `L1_4774:6292_INSERT_SIZE` = `L1_4774:6292`
    )

  dplyr::left_join(reads_wide, ins_wide, by = id_vars) %>%
    tibble::column_to_rownames(var = "sample_name")
}

# --- Build training matrix (Is_ctDNA label retained) ---
training_ <- build_smry_base(filter_to_training = TRUE) %>%
  pivot_to_matrix(has_label = TRUE) %>%
  dplyr::mutate(Is_ctDNA = factor(Is_ctDNA, levels = c("+ve", "-ve"), ordered = TRUE))

# --- Build test matrix (all HPV-16 samples, no label filter) ---
test_ <- build_smry_base(filter_to_training = FALSE) %>%
  pivot_to_matrix(has_label = FALSE)
```

```{r fit_lda_and_write, eval = TRUE, include = FALSE}

# --- Fit LDA on training set, predict on all HPV-16 samples ---
fit_ <- MASS::lda(Is_ctDNA ~ ., data = training_)
res_ <- predict(object = fit_, newdata = test_)

# --- Build output table and write to disk ---
posterior_probs <- test_ %>%
  tibble::rownames_to_column(var = "sample_name") %>%
  dplyr::mutate(
    Is_ctDNA  = res_$class,
    `Pr(x=0)` = res_$posterior[, "-ve"],
    `Pr(x=1)` = res_$posterior[, "+ve"]
  ) %>%
  dplyr::as_tibble() %>%
  readr::type_convert() %>%
  dplyr::left_join(mut_af_smry, by = "sample_name") %>%
  dplyr::left_join(manifest, by = "sample_name") %>%
  dplyr::mutate(index = as.integer(gsub("CTMS-", "", patient_id_mskcc, fixed = TRUE))) %>%
  dplyr::arrange(index) %>%
  dplyr::mutate(patient_id_mskcc = factor(patient_id_mskcc, levels = unique(patient_id_mskcc), ordered = TRUE)) %>%
  dplyr::select(-index)

readr::write_tsv(posterior_probs, file = url_output, append = FALSE, col_names = TRUE)
```

## Output summary

```{r summary_training, eval = TRUE, echo = FALSE}

n_pos <- sum(training_$Is_ctDNA == "+ve")
n_neg <- sum(training_$Is_ctDNA == "-ve")

cat(sprintf(
  "Training set: %d samples (%d ctDNA +ve, %d ctDNA -ve)\n",
  n_pos + n_neg, n_pos, n_neg
))
```

```{r summary_posterior, eval = TRUE, echo = FALSE}

class_smry <- posterior_probs %>%
  dplyr::count(Is_ctDNA, name = "n_samples") %>%
  dplyr::mutate(
    pct = round(100 * n_samples / sum(n_samples), 1)
  )

knitr::kable(
  class_smry,
  caption = "Classification summary across all HPV-16 samples",
  col.names = c("LDA Class", "N Samples", "% of Total")
)
```

```{r summary_posterior_probs, eval = TRUE, echo = FALSE}

prob_smry <- posterior_probs %>%
  dplyr::group_by(Is_ctDNA) %>%
  dplyr::summarize(
    `Mean Pr(+ve)` = round(mean(`Pr(x=1)`), 3),
    `SD Pr(+ve)`   = round(sd(`Pr(x=1)`),   3),
    `Min Pr(+ve)`  = round(min(`Pr(x=1)`),   3),
    `Max Pr(+ve)`  = round(max(`Pr(x=1)`),   3),
    .groups = "drop"
  )

knitr::kable(
  prob_smry,
  caption = "Posterior probability Pr(ctDNA = +ve) by predicted class"
)
```

```{r summary_af, eval = TRUE, echo = FALSE}

af_smry <- posterior_probs %>%
  dplyr::filter(!is.na(mean_af)) %>%
  dplyr::group_by(Is_ctDNA) %>%
  dplyr::summarize(
    `Mean AF`   = round(mean(mean_af) * 100, 3),
    `Median AF` = round(median(mean_af) * 100, 3),
    `Max AF`    = round(max(mean_af) * 100, 3),
    .groups = "drop"
  )

knitr::kable(
  af_smry,
  caption = "Mean allele fraction (%) by predicted class"
)
```

```{r summary_head, eval = TRUE, echo = FALSE}

knitr::kable(
  posterior_probs %>%
    dplyr::select(
      sample_name, patient_id_mskcc, Is_ctDNA,
      `Pr(x=0)`, `Pr(x=1)`, mean_af,
      timepoint_days_since_start_of_RT
    ) %>%
    dplyr::arrange(desc(`Pr(x=1)`)) %>%
    head(10),
  digits  = 3,
  caption = "Top 10 samples by posterior probability of ctDNA detection"
)
```

**Output written to:** `r url_output` (`r nrow(posterior_probs)` rows, `r ncol(posterior_probs)` columns)
